A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-06-04 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
Typos

\begin{abstract}
	We present a physically motivated generalization of probability theory that is suitable for classical mechanics, quantum mechanics and any future physical theory that allows a statistical description. The goal is to put the use of classical and quantum probability in a broader context, and show how the current mathematical structures are likely not suitable to solve the open problems in the foundations of physics. For the more mathematically inclined, we will point to areas where new math or generalization of established math are needed. For the more philosophically incline, we will point to areas where further conceptual work is needed.

@@

The above two approaches exhaust what can be done with additive measures, so we turn to the non-additive case. From the entropy, we can define the state capacity, which generalizes the notion of count of states. For each ensemble, we can define the fraction capacity, which generalizes the notion of probability. Both are non-negative sub-additive monotonic set functions, with the second being unit bounded. These measures are always unique and, since they are defined on the ensemble space itself, always have provide a physically meaningful representation. Moreover, in the compact case, these measure can be restricted on the extreme points while retaining uniqueness. (TODO: prove)

@@

Additional, each ensemble space $\Ens$ is equipped with a continuous entropy function $S : \Ens \to \mathbb{R}$. The entropy of a statistical mixture satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + (1-p) &S(\ens[b]) \leq S(p \ens[a] + (1-p) \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + (1-p) S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{In the larger work, we can show that the Shannon entropy is the only upper bound possible.} The bounds come from assuming that the entropy quantifies the variability of each ensemble. Intuitively, the variability cannot decrease during mixing, and will have an upper bound when mixing two ensembles that are non-overlapping, or orthogonal. In fact, we define two ensembles $\ens[a] \ortho \ens[b]$ to be orthogonal when they mixture gives a maximal entropy increase. This does recover the notion of orthogonality in the corresponding classical and quantum vector spaces of statistical ensembles. Lastly, mixtures preserve orthogonality in the sense that mixing non-overlapping ensembles still gives non-overlapping ensembles. More precisely, $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$.

@@

First, let us understand the physical significance. An ensemble $\ens$ can be seen as a probability measure $\mu$ over the ``pure states'' (i.e. the extreme points of $E$) if the value for every statistical quantity $f$ for $\ens$ can be recovered as an expectation over the pure states using the probability measure. That is, all statistical properties of $\ens$ can be recovered by the probability measure. This is what allows to think go $\ens$ as a probability distribution. Note, however, that the Choquet theorem does not guarantee uniqueness. The probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choque simplex by the property of having a single decomposition in terms of extreme points.

@@

We can apply Choquet theorem in the case where the closure of $\Ens$ in $V$ is a compact set $E$, which will also be convex\footnote{The closure of a convex set is convex. TODO: finalize the proof} and metrizable. In this case, we are guaranteed that all ensembles allows a representation through a probability on points that may corresponds idealized ensembles (i.e. the Dirac measure) that are not, even conceptually, realizable. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distribution over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems.

@@

Chouqet theory tells, in the realm where it can be applied, that classical probability fails in quantum mechanics precisely because the ensemble space of a quantum system allows multiple decomposition.

@@

Another way to represent quantum ensembles is through the use of quasiprobability distributions. Given a space $X$, a measure $\mu$ is affine if $\mu(X)$. All probability measures are affine measures, but the reverse is not true as affine measures can be negative on particular regions. The Wigner function, for example, can be negative over small regions of phase space.

@@

In general, since the ensemble space is bounded along every direction, not all affine measure will correspond to an ensemble space. Whether all probability measure correspond to a valid ensemble, or whether affine measures are even used, correspond to relationships between $\Ens$ and the space of probability measures $M_p(X)$. If $\Ens \subset M_p(X)$, then all states are represented by a probability measure. However, this is not a ``probability'' in the standard sense of the word. For example, the Husimi Q representation is always non-negative at every point, but two points to do not represent mutually exclusive events as they are not orthogonal (i.e. their inner product is non-zero and they do not maximize the entropy increase during mixture). Mathematically, the issue is that the points $X$ of the Husimi Q representation are not the extreme points of the closure ensembles space $\overline{\Ens}$. In general, $\Ens$ and $M_p(X)$ will overlap, but neither will contain the other. For example, in the Wigner representation the only pure states (i.e. the extreme points of $\Ens$) represented by a probability measure are the Gaussian wave-packets. This means that any other non-negative Wigner function will be in the convex hull of Gaussian wave-packets, and therefore will allow a decomposition in terms of them.

>>>>

\begin{abstract}
	We present a physically motivated generalization of probability theory that is suitable for classical mechanics, quantum mechanics and any future physical theory that allows a statistical description. The goal is to put the use of classical and quantum probability in a broader context, and show how the current mathematical structures are likely not suitable to solve the open problems in the foundations of physics. For the more mathematically inclined, we will point to areas where new math or generalization of established math are needed. For the more philosophically inclined, we will point to areas where further conceptual work is needed.

@@

The above two approaches exhaust what can be done with additive measures, so we turn to the non-additive case. From the entropy, we can define the state capacity, which generalizes the notion of count of states. For each ensemble, we can define the fraction capacity, which generalizes the notion of probability. Both are non-negative sub-additive monotonic set functions, with the second being unit bounded. These measures are always unique and, since they are defined on the ensemble space itself, always provide a physically meaningful representation. Moreover, in the compact case, these measures can be restricted to the extreme points while retaining uniqueness. (TODO: prove)

@@

Additionally, each ensemble space $\Ens$ is equipped with a continuous entropy function $S : \Ens \to \mathbb{R}$. The entropy of a statistical mixture satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + (1-p) &S(\ens[b]) \leq S(p \ens[a] + (1-p) \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + (1-p) S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{In the larger work, we can show that the Shannon entropy is the only upper bound possible.} The bounds come from assuming that the entropy quantifies the variability of each ensemble. Intuitively, the variability cannot decrease during mixing, and will have an upper bound when mixing two ensembles that are non-overlapping, or orthogonal. In fact, we define two ensembles $\ens[a] \ortho \ens[b]$ to be orthogonal when their mixture gives a maximal entropy increase. This does recover the notion of orthogonality in the corresponding classical and quantum vector spaces of statistical ensembles. Lastly, mixtures preserve orthogonality in the sense that mixing non-overlapping ensembles still gives non-overlapping ensembles. More precisely, $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$.

@@

First, let us understand the physical significance. An ensemble $\ens$ can be seen as a probability measure $\mu$ over the ``pure states'' (i.e. the extreme points of $E$) if the value for every statistical quantity $f$ for $\ens$ can be recovered as an expectation over the pure states using the probability measure. That is, all statistical properties of $\ens$ can be recovered by the probability measure. This is what allows us to think of $\ens$ as a probability distribution. Note, however, that the Choquet theorem does not guarantee uniqueness. The probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choque simplex by the property of having a single decomposition in terms of extreme points.

@@

We can apply Choquet's theorem in the case where the closure of $\Ens$ in $V$ is a compact set $E$, which will also be convex\footnote{The closure of a convex set is convex. TODO: finalize the proof} and metrizable. In this case, we are guaranteed that all ensembles allow a representation through a probability on points that may correspond to idealized ensembles (i.e. the Dirac measure) that are not, even conceptually, realizable. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distribution over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems.

@@

Chouqet theory tells us, in the realm where it can be applied, that classical probability fails in quantum mechanics precisely because the ensemble space of a quantum system allows multiple decompositions.

@@

Another way to represent quantum ensembles is through the use of quasiprobability distributions. Given a measurable space $(X, \Omega_X)$, a measure $\mu : \Omega_X \to \mathbb{R}$ is affine if $\mu(X) = 1$. All probability measures are affine measures, but the reverse is not true, as affine measures can be negative on particular regions. The Wigner function, for example, can be negative over small regions of phase space.

@@

In general, since the ensemble space is bounded along every direction, not all affine measures will correspond to an ensemble space. Whether all probability measures correspond to a valid ensemble, or whether affine measures are even used, corresponds to relationships between $\Ens$ and the space of probability measures $M_p(X)$. If $\Ens \subset M_p(X)$, then all states are represented by a probability measure. However, this is not a ``probability'' in the standard sense of the word. For example, the Husimi Q representation is always non-negative at every point, but two points do not represent mutually exclusive events as they are not orthogonal (i.e. their inner product is non-zero and they do not maximize the entropy increase during mixture). Mathematically, the issue is that the points $X$ of the Husimi Q representation are not the extreme points of the closure ensemble space $\overline{\Ens}$. In general, $\Ens$ and $M_p(X)$ will overlap, but neither will contain the other. For example, in the Wigner representation, the only pure states (i.e. the extreme points of $\Ens$) represented by a probability measure are the Gaussian wave-packets. This means that any other non-negative Wigner function will be in the convex hull of Gaussian wave-packets, and therefore will allow a decomposition in terms of them.
